### 2021-12-18 Manuscript.Rmd
manuscript edits, reviewer responses

---
bibliography: references.bib
csl: genomebiology.csl
site: "bookdown::bookdown_site"
output:
  bookdown::pdf_document2:
    toc: FALSE
    fig_caption: yes
---

@@

	# Remove tau for now since it is on a different scale
	edges_tidy<-edges_tidy %>% select(-tau)

@@

GO_by_tree<-strsplit(GO_to_tips$GO,";")
names(GO_by_tree)<-GO_to_tips$gene_tree

@@

@@

@@

**Background**

Siphonophores are complex colonial animals, consisting of asexually-produced bodies (called zooids) that are functionally specialized for specific tasks, including feeding, swimming, and sexual reproduction. Though this extreme functional specialization has captivated biologists for generations, its genomic underpinnings remain unknown. We use RNA-seq to investigate gene expression patterns in five zooids and one specialized tissue (pneumatophore) across seven siphonophore species. Analyses of gene expression across species present several challenges, including identification of comparable expression changes on gene trees with complex histories of speciation, duplication, and loss. Here, we conduct three analyses of expression. First, we examine gene expression within species. Then, we conduct classical analyses examining expression patterns between species.  Lastly, we introduce Speciation Branch Filtering, which allows us to examine the evolution of expression in a phylogenetic framework.

**Results**

Within and across species, we identified hundreds of zooid-specific and species-specific genes, as well as a number of putative transcription factors showing differential expression in particular zooids and developmental stages. We found that gene expression patterns tended to be largely consistent in zooids with the same function across species, but also some large lineage-specific shifts in gene expression.

Our findings show that patterns of gene expression have the potential to define zooids in colonial organisms. We also show that traditional analyses of the evolution of gene expression focus on the tips of gene phylogenies, identifying large-scale expression patterns that are zooid or species variable. The new explicit phylogenetic approach we propose here focuses on branches (not tips) offering a deeper evolutionary perspective into specific changes in gene expression within zooids along all branches of the gene (and species) trees.

## Background {-}

Colonial animals, consisting of genetically identical bodies that are physically and physiologically connected, can be found across the metazoan tree of life [@hiebert2020coloniality]. Functional specialization of such bodies has evolved multiple times within colonial animals, with siphonophores in particular showing the greatest diversity of functionally specialized bodies [@Beklemishev1969]. Siphonophores are highly complex, colonial “superorganisms” consisting of asexually produced bodies (termed zooids) that are homologous to solitary free-living polyps and medusae (the typical body forms in Cnidaria), but that share a common gastrovascular cavity [@dunn2006evolution;@mackie1963; @mackie1986aggregates;@mackie1987siphonophore;@totton1965synopsis]. The extreme specialization of siphonophore zooids has been of central interest to zoologists since the 19th century, in part because these zooids are highly interdependent [@Beklemishev1969; @mackie1963]. Siphonophore zooids have been likened to animal organs, with each functionally specialized zooid performing distinct roles within the colony [@mackie1963]. While this analogy works well for understanding the function of zooids within the colony as a whole, the analogy falls short in terms of explaining the evolutionary origin and development of these biological units: functionally specialized zooids are evolutionarily homologous to free living organisms, they are multicellular, possess distinct zooid specific cell types, and show regional subfunctionalization [@church2015histology;@Mackie1960;@totton1965synopsis]. While the developmental mechanisms generating zooids are very different in different clades of colonial animals [@carre1967; @carre1969etude; @carre1991complete; @dunn2006evolution; @siebert2015stem], the evolutionary processes acting on zooids may be similar to those acting on other modular biological units such as cell type, tissue, and organ [@hiebert2020coloniality]. The cellular and molecular processes underlying the patterning and molecular function of functionally specialized cnidarian zooids remain an open biological question. In recent years there has been a focus in particular on differential gene expression patterns found in different functionally specialized zooids [@sanders2014differential; @sanders2015interspecific; @Siebert:2011gv].

@@

Third, we use a novel phylogenetic approach to investigate the evolution of gene expression on gene trees with complex evolutionary histories. Most evolutionary studies of gene expression focus exclusively on comparing expression values of strict orthologs (i.e., gene lineages related only by speciation events) which are shared across all species. Analyses of strict orthologs are limited to a subset of genes that have no evidence of duplication, and we call this approach Speciation Tree Filtering (STF) [e.g @Brawand:2011du; @cardoso2019gene; @Levin:2016bp; @yang2013organ]. This is an all or nothing approach, where gene trees or subtrees (clades) are retained if they have one sequence for each species, and discarded if there are multiple sequences in any of the species. Because this is usually such a small fraction of genes, traditional STF analyses discard a large fraction of the data, including many gene families that may be of interest to the investigator.

Because the history of genes is characterized by more complex scenarios involving gene duplication and speciation, we introduce a new method that takes advantage of this rich history to examine the evolution of gene expression across more genes and species. We call this method Speciation Branch Filtering (SBF) (Fig. \@ref(fig:methods-comp)), as we map expression data to gene phylogenies and identify equivalent branches in the species tree which are descended from speciation events in order to make valid comparisons across species. While STF is a gene tree filtering approach, where entire trees or subtrees are discarded due to even a single duplication event, SBF is a branch filtering approach. This means that SBF retains many informative branches from trees that would be removed by STF, and preserves many more evolutionary comparisons for analyses. We use SBF in our explicit phylogenetic analyses across species.

Briefly, the principles of SBF are as follows: first we identify speciation and duplication events in a given gene tree, specifically labelling nodes that correspond to speciation events in the species tree (Fig. \@ref(fig:methods-comp), step 1). Next, we map gene expression values to the tips of the gene tree (Fig. \@ref(fig:methods-comp), step 2), and use phylogenetic methods to reconstruct expression values at the internal nodes (Fig. \@ref(fig:methods-comp), step 3). Expression values are mapped and reconstructed for each zooid/tissue separately, with the assumption that the structure is homologous across species. Then, we calculate scaled expression values across branches, by subtracting the expression value at the child node from the parent node, and dividing by branch length (branch lengths are calibrated against branches in the species tree) (Fig. \@ref(fig:methods-comp), step 4). Finally, we identify branches in the gene trees that correspond to equivalent branches in the species tree (hereafter species-equivalent branches) (Fig. \@ref(fig:methods-comp), step 5). Species-equivalent branches are gene tree branches with parent and child nodes that are both speciation events and that correspond to branches in the species tree. Each branch in the species tree is given a unique identifier (i.e., a letter) and the species-equivalent branches in gene trees are given the same identifier (Fig. \@ref(fig:methods-comp), step 5). This method enables the selection of branches within gene trees that are equivalent to branches within the species tree, and are thus comparable with one another across all gene trees. Unlike the STF approach, this approach considers equivalent branches that are descended from speciation events, but that have more complex evolutionary histories. For example, due to deeper gene duplication events, gene trees often contain multiple branches that correspond to the same branch in the species tree (Fig. \@ref(fig:methods-comp), step 5). Our method allows us to consider all of these branches.

Using the methods described above, combined with classical differential gene expression (DGE) analysis, we conducted DGE analyses among siphonophore zooids, and one specialized tissue (the pneumatophore, a gas filled float), in seven siphonophore species to address three questions. First, we analyzed differences in gene expression within species to identify patterns that define particular zooid types, and gain insight into the genes that may be playing a role in the structure, maintenance, and functioning of zooids. We also examined whether gene expression patterns could distinguish novel, distinct species-specific zooid types within particular species. Second, we compared gene expression between species, using STF followed by linear models to identify zooid- and species-variable genes. And third, we used phylogenetic methods and SBF to identify putative clade/lineage or zooid specific expression patterns.

@@

pca_data<-prcomp(t(pcatest))
pca_data_perc<-round(100*pca_data$sdev^2/sum(pca_data$sdev^2),1)
test_names<-colnames(pcatest) %>% str_split("_")
species<-lapply(test_names, `[[`, 1) %>% unlist() %>% str_replace("\\."," ")
treatment<-lapply(test_names, `[[`, 2) %>% unlist() 
#replicate<-lapply(test_names, `[[`, 3) %>% unlist()
df_pca_data <- data.frame(PC1 = pca_data$x[,1], PC2 = pca_data$x[,2], species=species, treatment=treatment)
plot_pca_ortholog<-df_pca_data %>% ggplot()+geom_point(aes(PC1,PC2, col=treatment, shape=species)) + 
  scale_shape_manual(values=c("Agalma elegans"=15,"Nanomia bijuga"=0,"Diphyes dispar"=8,"Physalia physalis"=10,"Frillagalma vityazi"=16,"Bargmannia elongata"=9, "Apolemia lanosa"=23)) +
  xlab(paste0("PC1: ",pca_data_perc[1],"% variance")) +  ylab(paste0("PC2:",pca_data_perc[2],"% variance")) +
  scale_color_manual(values=c("Gonfem"="#BFA7CE","Gonmal"="#F2B6BC","Necdev"="#B9C977","Palmat"="#F4DF1C","Gasdev"="#F78508", "Gasmat"="#F9C153", "Pne"="#8DCCE2"))

@@

pca_data2<-prcomp(t(pcatest2))
pca_data_perc2<-round(100*pca_data2$sdev^2/sum(pca_data2$sdev^2),1)
test_names2<-colnames(pcatest2) %>% str_split("_")
species2<-lapply(test_names2, `[[`, 1) %>% unlist() %>% str_replace("\\."," ")
treatment2<-lapply(test_names2, `[[`, 2) %>% unlist() 
#replicate<-lapply(test_names, `[[`, 3) %>% unlist()
df_pca_data2 <- data.frame(PC1 = pca_data2$x[,1], PC2 = pca_data2$x[,2], species=species2, treatment=treatment2)
plot_pca_ortholog2<-df_pca_data2 %>% ggplot()+geom_point(aes(PC1,PC2, col=treatment, shape=species)) + 
  scale_shape_manual(values=c("Agalma elegans"=15,"Nanomia bijuga"=0,"Frillagalma vityazi"=16,"Bargmannia elongata"=9)) +
  xlab(paste0("PC1: ",pca_data_perc[1],"% variance")) +  ylab(paste0("PC2:",pca_data_perc[2],"% variance")) +
  scale_color_manual(values=c("GasdevGasmat"="#F78508","NecdevGasmat"="#B9C977","PalmatGasmat"="#F4DF1C","PneGasmat"="#8DCCE2"))

@@

Most comparative studies of gene expression focus exclusively on strict 1:1 orthologs, requiring an assessment of orthology across all species. We call this type of analysis Speciation Tree Filtering (STF), as it limits analyses to a subset of genes with very specific evolutionary histories. Following this approach, we used Orthofinder 2 (v2.4.0) [@emms2019orthofinder] to identify strict orthologs. For all seven species, we identified `r Orthologs1$Orthogroup %>% unique() %>% length()` strict orthologs, of which `r test3 %>% nrow()` orthologs had expression data across all 7 species. In order to increase the number of genes for analysis, we focused on the four best sampled species (*Agalma elegans*, *Bargmannia elongata*, *Frillagalma vityazi*, and *Nanomia bijuga*). Using data from 4 taxa, we identified `r Orthologs2$Orthogroup %>% unique() %>% length()` strict orthologs, of which `r stvariance %>% nrow()` orthologs had expression values in four zooids/tissues: gastrozooids (developing and mature), nectophores (developing), palpons (mature), and the pneumatophore.  Using TPM10K alone, we found that orthologs clustered largely by species rather than zooid/ tissue (Fig. \@ref(fig:SOFT-plot)A). Following Breschi et al. @breschi2016gene (see methods), we used linear models to identify the proportion of variance that could be explained by species or zooid/tissue (Fig. \@ref(fig:SOFT-plot)B), and found that of `r stvariance  %>% nrow()` total orthologs, `r stvariance %>% filter(class=="SVG") %>% nrow()` species-variable genes (SVG), and `r stvariance %>% filter(class=="TVG") %>% nrow()` zooid/tissue-variable genes (TVG).

@@

The vast majority of gene trees contain species-equivalent branches with neutral changes, including a number of transcription factors and morphogenic signalling pathway genes. For example, in ratios of developing and mature gastrozooids, among the relevant species-equivalent branches in the Wnt gene phylogeny, changes are neutral across Wnt3 and Wnt2 branches, with the exception of a slight positive change across the branch leading to a *Diphyes dispar* Wnt2 paralog, suggesting a higher relative expression of this gene in the developing gastrozooid of *D. dispar* (Fig. \@ref(fig:wntplot)). For all zooid ratios, we find very consistent Wnt3 expression patterns with neutral or very small positive or negative changes in expression ratio across the branches. Indeed, across all cnidarians examined, Wnt3 has consistent localised expression at the oral pole, likely playing a role in axis formation and maintenance [@bagaeva2019wnt; @guder2006wnt; @hensel2014lineage; @hobmayer2000wnt; @momose2008maternally; @nawrocki2013expression; @sanders2015patterns]. However, for Pneumatophore/Gastrozooid ratios, some large changes are seen across branches K and L, leading to a *Agalma elegans* and *Nanomia bijuga* Wnt2 paralog respectively, likewise for Palpon/Gastrozooid a large change is seen across branch J, leading to a *Frillagalma vityazi* Wnt2 paralog. In *Nematostella vectensis* and *Hydractinia echinata*, Wnt2 is expressed in the middle of the polyp [@hensel2014lineage; @kusserow2005unexpected]. Without more detailed spatial expression patterns, it is difficult to know whether these differences in expression between species reflect species-specific differences in axial patterning and morphogenesis, such as an expansion or reduction of the expression domain.

@@

We also looked at gene trees with species-relative branches that have very large changes ( 5 < change < -5), representing putative lineage-specific expression patterns. In Frizzled5/8, Wnt4, and homeobox (putatively Hox-B8) gene-trees, for example, we found very large positive changes in ratios of developing gastroozoid/mature gastrozooids across the same species-equivalent branches, J (leading to *Frillagalma vityazi*) and D (leading to Euphysonectae, the clade comprising *A. elegans*, *N. bijuga*, and *F. vityazi*). In Frizzled5/8 for developing nectophores/gastrozooids we also see a large positive change across branch L (leading to *Nanomia bijuga*) and J (leading to *Frillagalma vityazi*) and negative change across branch K (leading to *Agalma elegans*). In ratios of developing nectophores/gastrozooids, we also found very large positive changes in branch J for Wnt-7b-like, Wnt-4, (putatively) Hox-B8 and NKX1-2. Likewise in pneumatophore/gastrozooid ratios, we also find large positive changes in branch J and L for Frizzled5/8, and a negative change across branch K. Across J, we also see positive changes in pneumatophore/gastrozooid in Hox-B8 and NKX1-2. Finally, for branch K, we see large positive changes in Wnt4 and Wnt7b-like. Many of these genes have been shown to have specific expression domains patterning cnidarian bodies [@hensel2014lineage; @kusserow2005unexpected; @ryan2007pre; @sinigaglia2013bilaterian], and further detailed expression analyses of these genes and others within the zooid are required to determine whether the patterns identified here reflect differences in expression domain between species and zooid.

@@

With our STF and linear model analyses we found that the vast majority of identified orthologs were neither exclusively species-variable or tissue-variable, however we were nevertheless able to identify a subset of either tissue- or species- variable genes. As has been observed for vertebrate organs, we find based on GO terms that identified species-variable genes tend to play a role in basic cellular functions (so-called 'housekeeping genes'), as compared to tissue-variable genes [@@breschi2016gene].

@@

Where traditional approaches identify strict ortholog genes *before* conducting analyses, SBF uses information from all gene copies within a gene tree regardless of their evolutionary history of duplication or speciation. By filtering our data by branches, we consider expression patterns at both the tips and internal nodes of gene phylogenies. We are thus able to identify shifts in expression leading to tips as well as shifts in expression leading to particular clades. Although we focus here on expression following speciation events, it is also possible to compare expression following duplication and speciation events.

@@

For Speciation Tree Filtering (STF) analyses, single copy orthologs were obtained using Orthofinder 2 (v2.4.0) [@emms2019orthofinder]. Linear models used in STF analyses were constructed using `lm()`, following methods and code developed by Breschi et al. 2016 [@breschi2016gene], https://github.com/abreschi/Rscripts/blob/master/anova.R.  All SVGs and TVGs are genes for which both species and tissue explain greater than 75% of variance. Additionally, in SVGs, the proportion of variance explained by species is two times greater than that explained by tissues; while in TVGs the proportion of variance explained by tissues is two times greater than that explained by species.

>>>>

---
bibliography: references.bib
csl: council-of-science-editors-author-date.csl
site: "bookdown::bookdown_site"
header-includes:
    - \usepackage{setspace}\doublespacing

output:
  bookdown::pdf_document2:
    toc: FALSE
    fig_caption: yes
---

@@

	# Remove tau for now since it is on a different scale
	edges_tidy<-edges_tidy %>% dplyr::select(-tau)

@@

GO_by_tree<-strsplit(GO_to_tips$GO,";")
names(GO_by_tree)<-GO_to_tips$gene_tree
gene_tree_names<-names(GO_by_tree)

fishertestSBF<-function(gene_tree, GO_ontology){

geneList <- factor(as.integer(gene_tree_names %in% unique(gene_tree)))
names(geneList) <- gene_tree_names

GOdata <- new("topGOdata", ontology = GO_ontology, allGenes = geneList, annot = annFUN.gene2GO, gene2GO = GO_by_tree)

@@

Article

@@

Keywords: Siphonophora, Cnidaria, expression evolution, functional specialization

@@

Siphonophores are complex colonial animals, consisting of asexually-produced bodies (zooids) that are functionally specialized for specific tasks, including feeding, swimming, and sexual reproduction. Though this extreme functional specialization has captivated biologists for generations, its genomic underpinnings remain unknown. We use RNA-seq to investigate gene expression patterns in five zooids and one specialized tissue across seven siphonophore species. Analyses of gene expression across species present several challenges, including identification of comparable expression changes on gene trees with complex histories of speciation, duplication, and loss. We examine gene expression within species, conduct classical analyses examining expression patterns between species, and introduce Speciation Branch Filtering, which allows us to examine the evolution of expression in a phylogenetic framework. Within and across species, we identified hundreds of zooid-specific and species-specific genes, as well as a number of putative transcription factors showing differential expression in particular zooids and developmental stages. We found that gene expression patterns tended to be largely consistent in zooids with the same function across species, but also some large lineage-specific shifts in gene expression. Our findings show that patterns of gene expression have the potential to define zooids in colonial organisms. Traditional analyses of the evolution of gene expression focus on the tips of gene phylogenies, identifying large-scale expression patterns that are zooid or species variable. The new explicit phylogenetic approach we propose here focuses on branches (not tips) offering a deeper evolutionary perspective into specific changes in gene expression within zooids along all branches of the gene (and species) trees.

## Introduction {-}

Colonial animals, consisting of genetically identical bodies that are physically and physiologically connected, can be found across the metazoan tree of life [@hiebert2020coloniality]. Functional specialization of such bodies has evolved multiple times within colonial animals, with siphonophores in particular showing the greatest diversity of functionally specialized bodies [@Beklemishev1969]. Siphonophores are highly complex, colonial “superorganisms” consisting of asexually produced bodies (termed zooids) that are homologous to solitary free-living polyps and medusae (the typical body forms in Cnidaria), but that share a common gastrovascular cavity [@dunn2006evolution;@mackie1963; @mackie1986aggregates;@mackie1987siphonophore;@totton1965synopsis]. The extreme specialization of siphonophore zooids has been of central interest to zoologists since the 19th century, in part because these zooids are highly interdependent [@Beklemishev1969; @mackie1963]. Siphonophore zooids have been likened to animal organs, with each functionally specialized zooid performing distinct roles within the colony [@mackie1963]. While this analogy works well for understanding the function of zooids within the colony as a whole, the analogy falls short in terms of explaining the evolutionary origin and development of these biological units: functionally specialized zooids are evolutionarily homologous to free living organisms, they are multicellular, possess distinct zooid specific cell types, and show regional subfunctionalization [@church2015histology; @Mackie1960; @totton1965synopsis]. While the developmental mechanisms generating zooids are very different in different clades of colonial animals [@carre1967; @carre1969etude; @carre1991complete; @dunn2006evolution; @siebert2015stem], the evolutionary processes acting on zooids may be similar to those acting on other modular biological units such as cell type, tissue, and organ [@hiebert2020coloniality]. The cellular and molecular processes underlying the patterning and molecular function of functionally specialized cnidarian zooids remain an open biological question. In recent years there has been a focus in particular on differential gene expression patterns found in different functionally specialized zooids [@10.1093/icb/icu080; @sanders2014differential; @sanders2015interspecific; @Siebert:2011gv].

@@

Third, we use a novel phylogenetic approach to investigate the evolution of gene expression on gene trees with complex evolutionary histories. Most evolutionary studies of gene expression focus exclusively on comparing expression values of strict orthologs (i.e., gene lineages related only by speciation events) which are shared across all species. Analyses of strict orthologs are limited to a subset of genes that have no evidence of duplication, and we call this approach Speciation Tree Filtering (STF) [e.g @Brawand:2011du; @cardoso2019gene; @Levin:2016bp; @yang2013organ]. Because the history of genes is characterized by more complex scenarios involving gene duplication and speciation, we introduce a new method that takes advantage of this rich history to examine the evolution of gene expression across more genes and species. We call this method Speciation Branch Filtering (SBF) (Fig. \@ref(fig:methods-comp)), as we map expression data to gene phylogenies and identify equivalent branches in the species tree which are descended from speciation events in order to make valid comparisons across species.

Using STF and SBF, combined with classical differential gene expression (DGE) analysis, we conducted DGE analyses among siphonophore zooids, and one specialized tissue (the pneumatophore, a gas filled float), in seven siphonophore species to address three questions. First, we analyzed differences in gene expression within species to identify patterns that define particular zooid types, and gain insight into the genes that may be playing a role in the structure, maintenance, and functioning of zooids. We also examined whether gene expression patterns could distinguish novel, distinct species-specific zooid types within particular species. Second, we compared gene expression between species, using STF followed by linear models to identify zooid- and species-variable genes. And third, we used phylogenetic methods and SBF to identify putative clade/lineage or zooid specific expression patterns.

@@

pca_data<-prcomp(t(pcatest))
pca_data_perc<-round(100*pca_data$sdev^2/sum(pca_data$sdev^2),1)
test_names<-colnames(pcatest) %>% str_split("_")
species<-lapply(test_names, `[[`, 1) %>% unlist() %>% str_replace("\\."," ")
treatment<-lapply(test_names, `[[`, 2) %>% unlist() 
#replicate<-lapply(test_names, `[[`, 3) %>% unlist()
df_pca_data <- data.frame(PC1 = pca_data$x[,1], PC2 = pca_data$x[,2], species=species, treatment=treatment)
plot_pca_ortholog<-df_pca_data %>% ggplot()+geom_point(aes(PC1,PC2, col=treatment, shape=species),size=3) + 
  scale_shape_manual(values=c("Agalma elegans"=15,"Nanomia bijuga"=0,"Diphyes dispar"=8,"Physalia physalis"=10,"Frillagalma vityazi"=16,"Bargmannia elongata"=9, "Apolemia lanosa"=23)) +
  xlab(paste0("PC1: ",pca_data_perc[1],"% variance")) +  ylab(paste0("PC2:",pca_data_perc[2],"% variance")) +
  scale_color_manual(values=c("Gonfem"="#BFA7CE","Gonmal"="#F2B6BC","Necdev"="#B9C977","Palmat"="#F4DF1C","Gasdev"="#F78508", "Gasmat"="#F9C153", "Pne"="#8DCCE2"))

@@

pca_data2<-prcomp(t(pcatest2))
pca_data_perc2<-round(100*pca_data2$sdev^2/sum(pca_data2$sdev^2),1)
test_names2<-colnames(pcatest2) %>% str_split("_")
species2<-lapply(test_names2, `[[`, 1) %>% unlist() %>% str_replace("\\."," ")
treatment2<-lapply(test_names2, `[[`, 2) %>% unlist() 
#replicate<-lapply(test_names, `[[`, 3) %>% unlist()
df_pca_data2 <- data.frame(PC1 = pca_data2$x[,1], PC2 = pca_data2$x[,2], species=species2, treatment=treatment2)
plot_pca_ortholog2<-df_pca_data2 %>% ggplot()+geom_point(aes(PC1,PC2, col=treatment, shape=species), size = 3) + 
  scale_shape_manual(values=c("Agalma elegans"=15,"Nanomia bijuga"=0,"Frillagalma vityazi"=16,"Bargmannia elongata"=9)) +
  xlab(paste0("PC1: ",pca_data_perc[1],"% variance")) +  ylab(paste0("PC2:",pca_data_perc[2],"% variance")) +
  scale_color_manual(values=c("GasdevGasmat"="#F78508","NecdevGasmat"="#B9C977","PalmatGasmat"="#F4DF1C","PneGasmat"="#8DCCE2"))

@@

Most comparative studies of gene expression focus exclusively on strict 1:1 orthologs, requiring an assessment of orthology across all species. We call this type of analysis Speciation Tree Filtering (STF), as it limits analyses to a subset of genes with very specific evolutionary histories. Following this approach, we used Orthofinder 2 (v2.4.0) [@emms2019orthofinder] to identify strict orthologs. For all seven species, we identified `r Orthologs1$Orthogroup %>% unique() %>% length()` strict orthologs, of which `r test3 %>% nrow()` orthologs had expression data across all 7 species. In order to increase the number of genes for analysis, we focused on the four best sampled species (*Agalma elegans*, *Bargmannia elongata*, *Frillagalma vityazi*, and *Nanomia bijuga*). Using data from 4 taxa, we identified `r Orthologs2$Orthogroup %>% unique() %>% length()` strict orthologs, of which `r stvariance %>% nrow()` orthologs had expression values in four zooids/tissues: gastrozooids (developing and mature), nectophores (developing), palpons (mature), and the pneumatophore.  Using TPM10K alone, we found that orthologs clustered largely by species rather than zooid/ tissue (Fig. \@ref(fig:SOFT-plot)A). Following @breschi2016gene (see methods), we used linear models to identify the proportion of variance that could be explained by species or zooid/tissue (Fig. \@ref(fig:SOFT-plot)B), and found that of `r stvariance  %>% nrow()` total orthologs, `r stvariance %>% filter(class=="SVG") %>% nrow()` species-variable genes (SVG), and `r stvariance %>% filter(class=="TVG") %>% nrow()` zooid/tissue-variable genes (TVG).

@@

The vast majority of gene trees contain species-equivalent branches with neutral changes, including a number of transcription factors and morphogenic signalling pathway genes. For example, in ratios of developing and mature gastrozooids, among the relevant species-equivalent branches in the Wnt gene phylogeny, changes are neutral across Wnt3 and Wnt2 branches, with the exception of a slight positive change across the branch leading to a *Diphyes dispar* Wnt2 paralog, suggesting a higher relative expression of this gene in the developing gastrozooid of *D. dispar* (Fig. \@ref(fig:wnts), \@ref(fig:wntplot)). For all zooid ratios, we find very consistent Wnt3 expression patterns with neutral or very small positive or negative changes in expression ratio across the branches. Indeed, across all cnidarians examined, Wnt3 has consistent localised expression at the oral pole, likely playing a role in axis formation and maintenance [@bagaeva2019wnt; @guder2006wnt; @hensel2014lineage; @hobmayer2000wnt; @momose2008maternally; @nawrocki2013expression; @sanders2015patterns]. However, for Pneumatophore/Gastrozooid ratios, some large changes are seen across branches K and L, leading to a *Agalma elegans* and *Nanomia bijuga* Wnt2 paralog respectively, likewise for Palpon/Gastrozooid a large change is seen across branch J, leading to a *Frillagalma vityazi* Wnt2 paralog. In *Nematostella vectensis* and *Hydractinia echinata*, Wnt2 is expressed in the middle of the polyp [@hensel2014lineage; @kusserow2005unexpected]. Without more detailed spatial expression patterns, it is difficult to know whether these differences in expression between species reflect species-specific differences in axial patterning and morphogenesis, such as an expansion or reduction of the expression domain.

@@

We also looked at the `r edges_tidy_directional %>% filter(change < -5 | change > 5) %>% .$gene_tree %>% unique() %>% length()` gene trees with species-relative branches that have very large changes ( 5 < change < -5), representing putative lineage-specific expression patterns. In Frizzled5/8, Wnt4, and homeobox (putatively Hox-B8) gene-trees, for example, we found very large positive changes in ratios of developing gastroozoid/mature gastrozooids across the same species-equivalent branches, J (leading to *Frillagalma vityazi*) and D (leading to Euphysonectae, the clade comprising *A. elegans*, *N. bijuga*, and *F. vityazi*). In Frizzled5/8 for developing nectophores/gastrozooids we also see a large positive change across branch L (leading to *Nanomia bijuga*) and J (leading to *Frillagalma vityazi*) and negative change across branch K (leading to *Agalma elegans*). In ratios of developing nectophores/gastrozooids, we also found very large positive changes in branch J for Wnt-7b-like, Wnt-4, (putatively) Hox-B8 and NKX1-2. Likewise in pneumatophore/gastrozooid ratios, we also find large positive changes in branch J and L for Frizzled5/8, and a negative change across branch K. Across J, we also see positive changes in pneumatophore/gastrozooid in Hox-B8 and NKX1-2. Finally, for branch K, we see large positive changes in Wnt4 and Wnt7b-like. Many of these genes have been shown to have specific expression domains patterning cnidarian bodies [@hensel2014lineage; @kusserow2005unexpected; @ryan2007pre; @sinigaglia2013bilaterian], and further detailed expression analyses of these genes and others within the zooid are required to determine whether the patterns identified here reflect differences in expression domain between species and zooid.

(ref:wnts) Pruned gene tree of Wnt genes in siphonophore species with expression values (TPM10K) at the tips and nodes. Developing nectophore/mature gastrozooid expression in green (above node), developing gastrozooid/mature gastrozooid expression in orange (below node). Changes across the branch are shown for developing gastrozooid/mature gastrozooid (orange). Red circles are duplication nodes and blue circles are speciation nodes.

```{r wnts, echo=FALSE, message=F, warning=FALSE, verbose= FALSE, fig.height=15, fig.width=15, fig.cap='(ref:wnts)', fig.align = 'center'}

###See sup figure code to see how these trees were generated.

knitr::include_graphics("Figures/Wntplot_new.pdf",auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE))

```{r GOterm enrichment analyses, results='asis', echo=FALSE, message=FALSE, warning=FALSE, verbose= FALSE}

#Positive GO term enrichment
#edges_tidy_directional %>% filter(change >2) %>% .$gene_tree %>% unique() %>% fishertestSBF(.,"BP")

#Negative GO term enrichment
#edges_tidy_directional %>% filter(change < -2) %>% .$gene_tree %>% unique() %>% fishertestSBF(.,"BP")

The nature of these analyses makes it difficult to investigate GO term enrichment, as each gene tree contains multiple genes often with diverse molecular function and that play diverse roles in biological processes -- additionally, this approach focuses on species-relative branches rather than tips. Nevertheless, we assigned GO terms at a gene tree level, and investigated GO term enrichment among gene trees that contain branches that have negative and positive changes. Among gene trees that contained species-relative branches with positive changes, we found an enrichment for a number of biological processes, including detection of stimulus, neuroblast proliferation, stem cell proliferation, and axon guidance. Among gene trees that contained species-relative branches with negative changes, we found in particular an enrichment for a number of metabolic processes including amine, NAD, tryptophan, and indolalkylamine metabolic process.

```{r BUSCO analyses, results='asis', echo=FALSE, message=FALSE, warning=FALSE, verbose= FALSE}

BUSCOall<-read_tsv("BUSCO/AllBUSCOs.tsv")

BUSCOall<-BUSCOall %>% tidyr::separate(col=Sequence,c("Species","gene"),sep="@", remove=FALSE) %>% mutate(Species=str_replace(Species,"_"," "))

sharedBUSCO<-BUSCOall %>% filter(Status=="Complete") %>% group_by(`Busco id`) %>% summarise(n=n())

##Number of BUSCO in expressiondataset
BUSCO_expression<-BUSCOall[BUSCOall$gene %in% Fulldataframe$sequence_ids,]

##Number of BUSCO in STF analyses
BUSCO_ortholog<-BUSCOall[BUSCOall$gene %in% Orthologs2$Gene,]

##Number of BUSCO in gene trees post filtering
BUSCO_genetree<-BUSCOall[BUSCOall$Sequence %in% nodes_ace$phy_node_names,]

terminaledges<-edges_tidy %>% filter(terminal==TRUE)
BUSCOnodes<-nodes_ace[nodes_ace$phy_node_names %in% BUSCOall$Sequence,]
tidyBUSCOs<-BUSCOnodes[BUSCOnodes$gene_tree %in% terminaledges$gene_tree,]

alignments<-read_tsv("Alignment_reports/All_expressiondata.tsv")

alignments<-alignments %>% mutate(`Percent aligned`=(`Reads aligned`/`Reads examined`)*100)

readr::write_csv(alignments, "Supplementary_Data/Supplementaryfile10.csv", na = "NA") 
alignments <- alignments %>% rename(Percent_aligned=`Percent aligned`)

#distribution of alignment quality
##alignments %>% ggplot()+geom_density(aes(Percent_aligned))

To assess the extent to which missing data may impact these analyses, we identified BUSCO scores for each of the reference transcriptomes (Table \@ref(tab:BUSCO-table)), finding BUSCO completeness ranging from 88.7% to 70.2% against the Metazoa BUSCO dataset [@10.1093/molbev/msab199]. Of this, `r round(nrow(BUSCO_expression)/nrow(BUSCOall)*100,2)`% of identified BUSCO genes are present in the initial expression dataset. In our strict ortholog STF analyses,  `r round(nrow(BUSCO_ortholog)/nrow(BUSCOall)*100,2)`% of identified BUSCO genes were retained. While in the SBF analyses, `r round(nrow(tidyBUSCOs)/nrow(BUSCOall)*100,2)`%  of identified BUSCO genes were retained.

The vast majority of zooid libraries mapped well to the reference transcriptomes, with the majority of libraries with between 85 and 95% of reads aligning to the reference, with an average of `r alignments$Percent_aligned %>% mean() %>% round(2)`% of reads aligned (supplementary file 10). Only eight libraries had poor alignment scores (<70%), notably all three developing nectophore libraries from *Nanomia bijuga* (56.53--63.08%), Male and female gonodendron from a single *Frillagalma vityazi* replicate (68.18% and 68.89% respectively), an *Apolemia lanosa* developing nectophore replicate (60.21%), and lastly a developing bract and pneumatophore replicate in *Bargmannia elongata* (64.46% and 49.61% respectively). The fact that all three *Nanomia bijuga* developing nectophore replicates had lower mapping scores, points to a reduced presence of likely nectophore-specific genes in the *Nanomia bijuga* reference transcriptome, which also had the lowest BUSCO score of all of the reference transcriptomes.

@@

With our STF and linear model analyses we found that the vast majority of identified orthologs were neither exclusively species-variable or tissue-variable, however we were nevertheless able to identify a subset of either tissue- or species- variable genes. As has been observed for vertebrate organs, we find based on GO terms that identified species-variable genes tend to play a role in basic cellular functions (so-called 'housekeeping genes'), as compared to tissue-variable genes [@breschi2016gene].

@@

Where traditional approaches identify strict ortholog genes *before* conducting analyses, SBF uses information from all gene copies within a gene tree regardless of their evolutionary history of duplication or speciation. STF is a gene tree filtering approach, where entire trees or subtrees are discarded due to even a single duplication event, SBF is a branch filtering approach. This means that SBF retains many informative branches from trees that would be removed by STF, and preserves many more evolutionary comparisons for analyses. By filtering our data by branches, we consider expression patterns at both the tips and internal nodes of gene phylogenies. We are thus able to identify shifts in expression leading to tips as well as shifts in expression leading to particular clades. Although we focus here on expression following speciation events, it is also possible to compare expression following duplication and speciation events.

@@

For Speciation Tree Filtering (STF) analyses, single copy orthologs were obtained using Orthofinder 2 (v2.4.0) [@emms2019orthofinder]. Linear models used in STF analyses were constructed using `lm()`, following methods and code developed by @breschi2016gene, https://github.com/abreschi/Rscripts/blob/master/anova.R.  All SVGs and TVGs are genes for which both species and tissue explain greater than 75% of variance. Additionally, in SVGs, the proportion of variance explained by species is two times greater than that explained by tissues; while in TVGs the proportion of variance explained by tissues is two times greater than that explained by species.
